---
title: "Data Manipulation with dplyr"
output:
  github_document: default
  html_notebook: default
---

Working with data using the [dplyr package](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html). 

### Load packages and the iris data set

```{r, warning=FALSE, message=FALSE}
library(dplyr); library(datasets)
data(iris)
```

### Glimpse at iris data
```{r}
glimpse(iris)

```
### select()
```{r, message=FALSE, warning=FALSE}
sepal_measurments <- select(iris, Sepal.Length, Sepal.Width)
glimpse(sepal_measurments)

```
#### select() with built-in filters
```{r}
petal_measurements <- select(iris, starts_with("Petal"))
glimpse(petal_measurements)
```
Other built-in column filters:  
1. `starts_with("X")`: every name that starts with "X"  
2. `ends_with("X")`: every name that ends with "X"    
3. `contains("X")`: every name that contains "X"  
4. `matches("X")`: every name that matches "X", where "X" can be a regular expression  
5. `num_range("x", 1:5)`: the variables named x01, x02, x03, x04 and x05  
6. `one_of(x)`: every name that appears in x, which should be a character vector.  

### mutate()
Add new columns based on existing
```{r, warning=FALSE}
df <- mutate(iris, Petal.Sum = Petal.Length + Petal.Width)
glimpse(df)
```
You can create more than one new column, even using newly created columns:
```{r, warning=FALSE}
df <- mutate(iris, Sepal.Sum = Sepal.Length + Sepal.Width, Petal.Sum = Petal.Length + Petal.Width, Mean.Sum = (Sepal.Sum + Petal.Sum)/2)
glimpse(df)
```


### filter()
Select rows based on logical criteria
```{r, message=FALSE, warning=FALSE}
filtered <- filter(iris, Sepal.Length > 5.1 | Sepal.Width > 3.5)
glimpse(filtered)
```

The usual logical operators are available inside `filter()`:  

* `x < y`  
* `x <= y`  
* `x == y`  
* `x != y`  
* `x >= y`  
* `x > y`  
* `x %in% c(a, b, c)` ,TRUE if `x` is in the vector `c(a, b, c)`  
* `is.na()`  

Similarly, the usual boolean operators work inside filter():  
* `&` (and)  
* `|` (or)  
* `!` (not)  

Note `&` can be replaced by commas. The following are equivalent:  
`filter(df, a > 0 & b > 0)`  
`filter(df, a > 0, b > 0)`  

### summarise()
Apply functions to columns of a dataframe.
```{r}
summarise(iris, mean_petal_length = mean(Petal.Length), sd_petal_length = sd(Petal.Length))

```


### arrange()
Sorts a dataframe on one or more column, by defualt in ascending order. To sort from highest to lowest, wrap the variable of interest in `desc()`.
```{r}
sorted <- arrange(iris, desc(Sepal.Length))
head(sorted)
```
Break ties with a second variable:
```{r}
sorted2 <- arrange(iris, desc(Sepal.Length), desc(Petal.Width))
head(sorted2)
```

### Pipe Operator
The pipe operator, `%>%` enables the chaining together of functions by supplying the output of one function as the first argument to another. It makes code more intuitive to both read and write.
```{r}
iris %>%
  filter(Sepal.Length > 7) %>%
  mutate(Approx.Petal.Area = Petal.Length * Petal.Width) %>%
  summarise(mean = mean(Approx.Petal.Area),
            sd = sd(Approx.Petal.Area), 
            max = max(Approx.Petal.Area),
            min = min(Approx.Petal.Area))
```


### group_by



### vectorized dplyr






